---
title: 'Week 15 - What is Statistical Data Mining: An Introductory Review of Methods'
date: "`r Sys.Date()`"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

---

## Learning objectives

By the end of this notebook, you will be able to:

> **Define** statistical data mining and distinguish it from traditional statistics and machine learning.

> **Apply** classification with a decision tree in R — build, visualise, and evaluate the model using a confusion matrix and accuracy.

> **Apply** k-means clustering in R — fit clusters, visualise results, and compare recovered groups to known labels.

> **Explain** the three key metrics — support, confidence, and lift — used in association rule mining, and mine rules from a transactional dataset.

> **Describe** the main techniques in text mining, time series analysis, and social network analysis, and run a worked example for each.

> **Identify** the current landscape of big data tools available to R users and explain when each is appropriate.

---

## Packages used in this notebook

```{r packages}
# --- Core data wrangling and visualisation ---
library(tidyverse)    # dplyr, ggplot2, tidyr, purrr, etc.

# --- Modelling framework ---
library(tidymodels)   # unified interface for train/test splits and evaluation

# --- Classification ---
library(rpart)        # CART decision trees (rpart)
library(partykit)     # conditional inference trees (ctree) and forests (cforest)
library(ranger)       # fast random forest implementation

# --- Association rule mining ---
library(arules)       # apriori() and eclat() for rule mining
library(arulesViz)    # visual exploration of association rules

# --- Text mining ---
library(tm)           # Corpus, DocumentTermMatrix, text cleaning pipeline
library(tidytext)     # tidy interface to text analysis (unnest_tokens, tf_idf)
library(wordcloud)    # word cloud visualisations

# --- Time series ---
library(forecast)     # auto.arima(), forecast(), autoplot() for time series

# --- Social network analysis ---
library(igraph)       # graph objects, centrality measures, community detection
```

---

## 1. What is statistical data mining?

**Statistical data mining** is the process of discovering patterns, relationships, and structure in large or complex datasets by combining statistical inference with computational algorithms. It is sometimes called **knowledge discovery in databases (KDD)**.

A few key distinctions:

| Concept | Focus | Key question |
|---|---|---|
| **Traditional statistics** | Testing hypotheses about small, clean datasets | Is this effect real? |
| **Machine learning** | Optimising predictive accuracy | Can I predict this accurately? |
| **Statistical data mining** | Discovering unknown structure in large, messy data | What patterns are hiding in this data? |

In practice the three overlap heavily. A data mining project might use statistical tests to validate patterns, machine learning algorithms to classify records, and visualisation to communicate structure.

### 1.1 The seven methods covered this week

The table below provides an orientation map for the whole notebook.

| Method | Supervised? | Output | Real-world example |
|---|---|---|---|
| **Classification** | Yes (labelled data) | A class label for each observation | Email spam / not spam |
| **Clustering** | No | Group memberships | Customer segments by purchase behaviour |
| **Association rules** | No | If-then rules between items | "Customers who buy bread also buy butter" |
| **Text mining** | No / semi | Word frequencies, topics, sentiment | Summarising 10,000 customer reviews |
| **Time series analysis** | No | Trend, seasonality, forecast | Predicting next quarter's sales |
| **Social network analysis** | No | Graph metrics, communities | Finding influential users in a network |
| **Big data** | Both | Scalable version of any of the above | Processing billions of web clicks |

---

## 2. Classification with R

### 2.1 What is classification?

**Classification** is the task of assigning each observation to one of a fixed set of pre-defined categories (classes), learning the decision rule from **labelled** training data. Because labels are provided, classification is **supervised learning**.

Key terms:

| Term | Definition |
|---|---|
| **Training set** | Labelled data used to fit the model |
| **Test set** | Held-out data used to evaluate performance honestly |
| **Decision boundary** | The line (or surface) separating predicted classes |
| **Confusion matrix** | A table comparing predicted vs. actual labels |
| **Accuracy** | Proportion of observations correctly classified |

Classification differs from **clustering** (no labels — groups are discovered, not assigned) and from **regression** (the output is a continuous number, not a category).

### 2.2 R packages for classification

| Package | Method | Notes |
|---|---|---|
| `partykit` | Conditional inference trees (`ctree`), random forests (`cforest`) | Statistically motivated splits; good default |
| `rpart` | CART decision trees | Classic algorithm; very widely used |
| `ranger` | Fast random forest | Best-in-class speed; recommended for production |
| `e1071` / `kernlab` | Support vector machines (SVM) | Strong on medium-sized datasets |
| `nnet` / `neuralnet` | Neural networks | Useful for simple architectures |
| `yardstick` (tidymodels) | Performance evaluation | Accuracy, ROC-AUC, confusion matrix |

### 2.3 Example: classifying iris flower species

The classic `iris` dataset contains measurements of 150 flowers across three species (*setosa*, *versicolor*, *virginica*). We will build and evaluate two classifiers: a conditional inference tree and a random forest.

#### Step 1 — Data preparation and train/test split

```{r classification_data}
# inspect the iris dataset structure
str(iris)

# set a random seed so results are reproducible
set.seed(465)

n <- nrow(iris)   # 150 observations

# randomly sample 30% of row indices for the test set
# the remaining 70% form the training set
ind <- sample.int(n, size = round(0.3 * n))

iris.train <- iris[-ind, ]   # training set: 70% of rows
iris.test  <- iris[ ind, ]   # test set:     30% of rows

cat("Training rows:", nrow(iris.train), "\n")
cat("Testing rows: ", nrow(iris.test),  "\n")

# baseline (null model) accuracy: always predict the majority class
# a good model should clearly beat this
majority_class <- names(sort(table(iris.train$Species), decreasing = TRUE))[1]
null_accuracy  <- mean(iris.test$Species == majority_class)
cat("Null model accuracy (always predict '", majority_class, "'):", round(null_accuracy, 3), "\n")
```

#### Step 2 — Build a conditional inference tree

A **conditional inference tree** (`ctree`) differs from a standard CART decision tree in how it chooses splits. Rather than maximising an impurity measure (like Gini index), `ctree` uses statistical significance tests at each node. This makes it less prone to overfitting and avoids the need to prune.

```{r classification_ctree}
# define the formula: predict Species from all four measurements
iris.formula <- Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width

# fit the conditional inference tree on training data only
iris.ctree <- partykit::ctree(iris.formula, data = iris.train)

# print a text summary of the tree structure
print(iris.ctree)
```

```{r classification_ctree_plot, fig.height=8, fig.width=10}
# visualise the tree
# each terminal node shows the predicted class distribution
plot(iris.ctree,
     main = "Conditional inference tree for iris species classification")
```

**Reading the tree:** Each internal node shows a variable and a split threshold chosen by a significance test. The terminal (leaf) nodes show bar charts of the class distribution for observations that reach that node. The predicted class for a new observation is the majority class in the leaf it falls into.

#### Step 3 — Predict and evaluate

```{r classification_predict}
# generate predicted class labels on the held-out test set
pred <- predict(iris.ctree, newdata = iris.test)

# confusion matrix: rows = actual class, columns = predicted class
cm <- table(Actual = iris.test$Species, Predicted = pred)
cm
```

**Reading the confusion matrix:**

- **Diagonal cells** (top-left to bottom-right) are correct predictions.
- **Off-diagonal cells** are errors. For example, if row = *versicolor* and column = *virginica*, those are versicolor flowers that the model incorrectly called virginica.
- A perfect classifier would have zeros everywhere off the diagonal.

```{r classification_accuracy}
# overall accuracy: sum of diagonal divided by total test observations
accuracy_ctree <- sum(diag(cm)) / sum(cm)
cat("Conditional inference tree accuracy:", round(accuracy_ctree, 3), "\n")
cat("Null model accuracy:                ", round(null_accuracy, 3), "\n")
cat("Improvement over null model:        ", round(accuracy_ctree - null_accuracy, 3), "\n")
```

#### Step 4 — Visualise the decision regions

The tree splits primarily on `Petal.Length` and `Petal.Width`. Plotting those two dimensions coloured by predicted class makes the decision boundaries intuitive.

```{r classification_regions, fig.height=5, fig.width=7}
# create a fine grid of petal measurements covering the observed range
grid <- expand.grid(
  Petal.Length = seq(min(iris$Petal.Length), max(iris$Petal.Length), length = 200),
  Petal.Width  = seq(min(iris$Petal.Width),  max(iris$Petal.Width),  length = 200),
  Sepal.Length = mean(iris$Sepal.Length),   # hold sepal measurements at their mean
  Sepal.Width  = mean(iris$Sepal.Width)
)

# predict species for every point on the grid
grid$Predicted <- predict(iris.ctree, newdata = grid)

# plot the decision regions as a background, then overlay real test observations
ggplot() +
  geom_tile(data = grid,
            aes(x = Petal.Length, y = Petal.Width, fill = Predicted),
            alpha = 0.3) +
  geom_point(data = iris.test,
             aes(x = Petal.Length, y = Petal.Width,
                 colour = Species, shape = Species),
             size = 3) +
  scale_fill_brewer(palette = "Set2")   +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title    = "Decision regions of the conditional inference tree",
    subtitle = "Background shading = predicted class; points = true species (test set)",
    x        = "Petal length (cm)",
    y        = "Petal width (cm)"
  ) +
  theme_minimal()
```

#### Step 5 — Random forest for comparison

A **random forest** fits many decision trees, each on a random bootstrap sample of the data and a random subset of features, then takes a majority vote. It almost always outperforms a single tree.

```{r classification_rf}
library(ranger)

# fit a random forest with 500 trees
rf_fit <- ranger(
  Species ~ .,
  data        = iris.train,
  num.trees   = 500,
  probability = FALSE,   # classification, not probability estimation
  seed        = 42
)

# predict on the test set
rf_pred     <- predict(rf_fit, data = iris.test)$predictions
rf_cm       <- table(Actual = iris.test$Species, Predicted = rf_pred)
rf_accuracy <- sum(diag(rf_cm)) / sum(rf_cm)

cat("Random forest accuracy:              ", round(rf_accuracy, 3), "\n")
cat("Conditional inference tree accuracy: ", round(accuracy_ctree, 3), "\n")
```

**Take-away:** The random forest typically achieves higher accuracy on the test set because averaging over many trees reduces variance. On a clean, small dataset like `iris` the difference is small — on noisier, larger data it is usually more substantial.

---

## 3. Clustering with R

### 3.1 What is clustering?

**Clustering** is an **unsupervised** method that groups observations so that members of the same cluster are more similar to each other than to members of other clusters. Unlike classification, there are no pre-existing labels — the algorithm discovers the groupings from the data itself.

Key terms:

| Term | Definition |
|---|---|
| **Centroid** | The mean position of all observations in a cluster — the cluster's "centre of gravity" |
| **Within-cluster SS (WCSS)** | The total squared distance of every observation from its cluster centroid. k-means minimises this. |
| **K** | The number of clusters, which must be specified before running k-means |
| **Inertia** | Another name for WCSS — lower inertia means tighter clusters |

### 3.2 R packages for clustering

| Package | Method |
|---|---|
| `stats` (base R) | `kmeans()` — k-means, `hclust()` — hierarchical clustering |
| `cluster` | `pam()` (k-medoids), `agnes()` (agglomerative), `diana()` (divisive) |
| `fpc` | `dbscan()` — density-based clustering, good for irregular shapes |
| `NbClust` | 30 indices for determining the optimal number of clusters |

### 3.3 Example: k-means clustering on iris

We remove the species labels, cluster purely on measurements, then compare the recovered clusters to the true species. This demonstrates both the power and the limits of k-means.

```{r clustering_kmeans}
set.seed(654)

# remove the species label — k-means sees only numeric measurements
iris2 <- iris %>% dplyr::select(-Species)

# fit k-means with k = 3 (we know there are 3 species, but the algorithm does not)
# nstart = 25: try 25 random initialisations and keep the best solution
iris.kmeans <- kmeans(iris2, centers = 3, nstart = 25)

# cluster sizes
cat("Cluster sizes:", iris.kmeans$size, "\n\n")

# cluster centroids (mean of each variable within each cluster)
cat("Cluster centroids:\n")
round(iris.kmeans$centers, 2)
```

```{r clustering_compare}
# compare recovered clusters to the true species labels
# rows = true species, columns = assigned cluster
cm_clust <- table(True_species = iris$Species, Cluster = iris.kmeans$cluster)
cm_clust

# how well did k-means recover the true groupings?
# find the best cluster-to-species assignment and compute overall agreement
best_match <- apply(cm_clust, 1, max)
agreement  <- sum(best_match) / nrow(iris)
cat("\nBest-case cluster-to-species agreement:", round(agreement, 3), "\n")
cat("(1.0 = perfect recovery; 0.333 = no better than random for 3 equal groups)\n")
```

**Interpreting the table:** If k-means perfectly recovered the species structure, each row would have a single non-zero entry. The actual table shows that *setosa* is almost always recovered cleanly (its petals are much smaller), while *versicolor* and *virginica* overlap more in measurement space.

```{r clustering_plot, fig.height=5, fig.width=8}
# plot the clusters with ggplot2, using Sepal.Length vs Sepal.Width
# (the two features used in the original slides — useful for comparison)

# attach cluster and true species labels for plotting
iris_plot <- iris %>%
  mutate(cluster = factor(iris.kmeans$cluster))

# extract centroids for annotation
centers_df <- as.data.frame(iris.kmeans$centers) %>%
  mutate(cluster = factor(1:3))

ggplot(iris_plot, aes(x = Sepal.Length, y = Sepal.Width)) +
  # colour by cluster assignment, shape by true species
  geom_point(aes(colour = cluster, shape = Species), size = 2.5, alpha = 0.8) +
  # mark centroids with a large X
  geom_point(data = centers_df,
             aes(x = Sepal.Length, y = Sepal.Width, colour = cluster),
             shape = 4, size = 6, stroke = 2) +
  scale_colour_brewer(palette = "Dark2", name = "k-means\ncluster") +
  scale_shape_manual(values = c(16, 17, 15), name = "True\nspecies") +
  labs(
    title    = "k-means clustering of iris (k = 3) on sepal dimensions",
    subtitle = "Colour = k-means cluster; shape = true species; X = cluster centroid",
    x        = "Sepal length (cm)",
    y        = "Sepal width (cm)"
  ) +
  theme_minimal()
```

**What the plot reveals:** The cluster boundaries in sepal space are less clean than in petal space. This is a useful reminder that k-means performance depends strongly on which features you include — sepal measurements are less discriminating than petal measurements for these three species.

---

## 4. Association rule mining with R

### 4.1 What is association rule mining?

**Association rule mining** finds frequent patterns, co-occurrences, and correlations in large transactional datasets. Each **transaction** is a set of items (e.g., a shopping basket). A **rule** has the form:

$$\{A, B\} \Rightarrow \{C\}$$

*"Customers who buy A and B also tend to buy C."*

Three metrics determine how interesting a rule is:

| Metric | Formula | Plain-English meaning |
|---|---|---|
| **Support** | $\frac{\text{transactions containing } A \cup C}{\text{all transactions}}$ | How common is this combination overall? |
| **Confidence** | $\frac{\text{transactions containing } A \cup C}{\text{transactions containing } A}$ | Among baskets with A, what fraction also contain C? |
| **Lift** | $\frac{\text{confidence}(A \Rightarrow C)}{\text{support}(C)}$ | How much more likely is C given A, compared to C alone? |

**Interpreting lift:**

- Lift = 1 → A and C are statistically independent (the rule adds no information)
- Lift > 1 → A and C co-occur more often than chance (a genuinely interesting rule)
- Lift < 1 → A and C are negatively associated (buying A makes C less likely)

### 4.2 R packages for association rules

| Package | Purpose |
|---|---|
| `arules` | `apriori()`, `eclat()` — mine rules; `inspect()` — display rules |
| `arulesSequence` | Sequential pattern mining (order of items matters) |
| `arulesViz` | Network graphs, scatter plots, grouped matrix visualisations of rules |

### 4.3 Example: grocery basket analysis

The `Groceries` dataset in the `arules` package contains 9,835 real grocery transactions. We mine association rules and identify the most interesting ones by lift.

```{r assoc_data}
data(Groceries)   # load from the arules package

# inspect the dataset structure
cat("Number of transactions:", length(Groceries), "\n")
cat("Number of items:       ", nitems(Groceries), "\n\n")

# most frequently purchased items
itemFrequencyPlot(Groceries,
                  topN  = 15,
                  type  = "relative",
                  col   = "steelblue",
                  main  = "Top 15 most frequently purchased grocery items",
                  ylab  = "Item frequency (proportion of transactions)")
```

```{r assoc_rules}
# mine association rules using the Apriori algorithm
# supp = 0.01: rule must appear in at least 1% of transactions
# conf = 0.40: at least 40% of baskets containing the LHS also contain the RHS
rules <- apriori(
  Groceries,
  parameter = list(supp = 0.01, conf = 0.40, minlen = 2)
)

cat("Number of rules found:", length(rules), "\n\n")

# display the top 10 rules sorted by lift (most interesting first)
cat("Top 10 rules by lift:\n")
arules::inspect(sort(rules, by = "lift")[1:10])
```

```{r assoc_viz, fig.height=6, fig.width=8}
# scatter plot: support vs confidence, coloured by lift
# rules in the top-right with high lift are the most actionable
top_rules <- sort(rules, by = "lift")[1:50]

plot(top_rules,
     measure  = c("support", "confidence"),
     shading  = "lift",
     main     = "Top 50 association rules: support vs confidence (colour = lift)",
     col      = colorRampPalette(c("steelblue", "tomato"))(100))
```

**Reading the scatter plot:** Each point is one rule. Rules to the upper right have both high support (common) and high confidence (reliable). The colour encodes lift — red points are the most interesting because the co-occurrence is far above what random chance would predict.

```{r assoc_network, fig.height=6, fig.width=8}
# network graph of the top 20 rules by lift
# nodes = items, edges = rules, edge width ∝ lift
plot(sort(rules, by = "lift")[1:20],
     method = "graph",
     control = list(type = "items"),
     main   = "Network of top 20 association rules (edge width ∝ lift)")
```

---

## 5. Text mining with R

### 5.1 What is text mining?

**Text mining** (also called **natural language processing** in machine learning contexts) is the process of extracting structured, quantitative information from unstructured text. Raw text cannot be fed directly into most statistical models — it must first be transformed into a numeric representation.

Key concepts:

| Concept | Definition |
|---|---|
| **Corpus** | The full collection of text documents being analysed |
| **Token** | A unit of text — typically a word, but can be a character, sentence, or n-gram |
| **Tokenisation** | Splitting a document into individual tokens |
| **Stop words** | Common words ("the", "and", "a") that carry little meaning and are usually removed |
| **Stemming / Lemmatisation** | Reducing words to their root form ("running" → "run") to merge related forms |
| **Bag of words** | Representing each document as a vector of word counts, ignoring word order |
| **TF-IDF** | Term frequency–inverse document frequency: weights words by how distinctive they are to a specific document |

**TF-IDF formula:**

$$\text{TF-IDF}(t, d) = \underbrace{\frac{\text{count of } t \text{ in } d}{\text{total words in } d}}_{\text{term frequency}} \times \underbrace{\log\left(\frac{\text{total documents}}{\text{documents containing } t}\right)}_{\text{inverse document frequency}}$$

A word common in one document but rare across the corpus gets a high TF-IDF score — it is likely a defining term for that document.

### 5.2 R packages for text mining

| Package | Purpose |
|---|---|
| `tm` | Corpus management, cleaning pipeline, `DocumentTermMatrix` |
| `tidytext` | Tidy interface: `unnest_tokens()`, `bind_tf_idf()`, integrates with `dplyr` |
| `topicmodels` | Latent Dirichlet Allocation (LDA) topic modelling |
| `wordcloud` | Word cloud visualisations |
| `stringr` | String manipulation (part of tidyverse) |

### 5.3 Example: analysing six short documents

We build a small corpus of six documents, clean it, and compute word frequencies and TF-IDF scores using both the `tm` and `tidytext` approaches.

```{r text_corpus}
# six short documents on different topics
docs <- c(
  doc1 = "Statistics is the science of learning from data. Data analysis and probability are central to statistics.",
  doc2 = "Machine learning algorithms learn patterns from training data to make predictions on new data.",
  doc3 = "Data science combines statistics, machine learning, and programming to extract insights from large datasets.",
  doc4 = "R is a statistical programming language widely used for data analysis, visualisation, and machine learning.",
  doc5 = "Neural networks are a class of machine learning models inspired by the structure of the human brain.",
  doc6 = "Probability theory provides the mathematical foundation for both statistics and machine learning."
)

# build a tm Corpus from the character vector
corpus <- tm::Corpus(VectorSource(docs))

# text cleaning pipeline
corpus_clean <- corpus %>%
  tm_map(content_transformer(tolower)) %>%   # lowercase everything
  tm_map(removePunctuation)                 %>%   # remove . , ! etc.
  tm_map(removeNumbers)                     %>%   # remove digits
  tm_map(removeWords, stopwords("english")) %>%   # remove "the", "and", etc.
  tm_map(stripWhitespace)                        # collapse multiple spaces

# inspect one cleaned document
cat("Cleaned document 1:\n", as.character(corpus_clean[[1]]), "\n")
```

```{r text_dtm}
# build a Document-Term Matrix (DTM)
# rows = documents, columns = unique words, cells = word count
dtm <- DocumentTermMatrix(corpus_clean)

cat("DTM dimensions:", nrow(dtm), "documents ×", ncol(dtm), "terms\n\n")

# find terms appearing in at least 3 documents
frequent_terms <- findFreqTerms(dtm, lowfreq = 3)
cat("Terms appearing in ≥ 3 documents:\n")
print(frequent_terms)
```

```{r text_wordcloud, fig.height=5, fig.width=6}
# compute overall word frequencies across all documents
word_freqs <- colSums(as.matrix(dtm))
word_freqs <- sort(word_freqs, decreasing = TRUE)

# word cloud: size ∝ frequency
set.seed(42)
wordcloud::wordcloud(
  words     = names(word_freqs),
  freq      = word_freqs,
  min.freq  = 1,
  max.words = 40,
  colors    = RColorBrewer::brewer.pal(8, "Dark2"),
  scale     = c(3, 0.5)
)
title("Word cloud: six data science documents")
```

```{r text_tfidf}
# TF-IDF using tidytext — integrates naturally with dplyr
library(tidytext)

# create a tidy data frame: one row per word per document
tidy_docs <- tibble(
  document = names(docs),
  text     = unname(docs)
) %>%
  unnest_tokens(word, text) %>%              # tokenise
  anti_join(stop_words, by = "word") %>%    # remove stop words
  count(document, word, sort = TRUE)        # count word occurrences

# compute TF-IDF scores
tidy_tfidf <- tidy_docs %>%
  bind_tf_idf(word, document, n)

# top 3 most distinctive words per document
tidy_tfidf %>%
  group_by(document) %>%
  slice_max(tf_idf, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(document, desc(tf_idf)) %>%
  dplyr::select(document, word, tf_idf) %>%
  mutate(tf_idf = round(tf_idf, 4))
```

**Interpreting TF-IDF:** Words like "statistics" score high in documents that use the word a lot (doc1, doc3, doc6) but not in others. Generic words like "data" appear everywhere and therefore have a lower IDF component — they do not distinguish one document from another.

---

## 6. Time series analysis with R

### 6.1 What is time series analysis?

A **time series** is a sequence of observations recorded at regular time intervals — hourly temperatures, monthly sales, annual GDP. Time series data violate the standard statistical assumption that observations are independent: values close in time tend to be similar.

Key concepts:

| Concept | Definition |
|---|---|
| **Trend** | A long-run increase or decrease in the level of the series |
| **Seasonality** | A repeating pattern at a fixed, known period (monthly, quarterly, annual) |
| **Residual / noise** | The remaining variation after trend and seasonality are removed |
| **Stationarity** | A series whose statistical properties (mean, variance) do not change over time. Required by many models. |
| **ARIMA** | AutoRegressive Integrated Moving Average — a flexible class of models for stationary time series |
| **Decomposition** | Separating a series into trend + seasonal + residual components |

### 6.2 R packages for time series

| Package | Function |
|---|---|
| `stats` (base R) | `decompose()`, `stl()` — additive/multiplicative decomposition; `arima()` — manual ARIMA |
| `forecast` | `auto.arima()` — automatic model selection; `forecast()` — generate predictions; `autoplot()` — ggplot2-compatible plots |
| `TSclust` | Time series clustering |
| `dtw` | Dynamic time warping — comparing series of different lengths or phases |

### 6.3 Example: airline passenger data

`AirPassengers` is a classic built-in R dataset: monthly totals of international airline passengers from 1949 to 1960 (in thousands). It is an ideal illustration because it has a clear upward trend and strong annual seasonality.

```{r ts_plot, fig.height=4, fig.width=8}
# AirPassengers is built into base R — no download needed
data(AirPassengers)

# basic plot of the raw series
autoplot(AirPassengers) +
  labs(
    title = "Monthly international airline passengers (1949–1960)",
    x     = "Year",
    y     = "Passengers (thousands)"
  ) +
  theme_minimal()
```

```{r ts_decompose, fig.height=6, fig.width=8}
# decompose into trend + seasonal + residual components
# multiplicative: seasonal variation grows with the level (appropriate here)
decomp <-  stats::decompose(AirPassengers, type = "multiplicative")

autoplot(decomp) +
  labs(title = "Multiplicative decomposition of AirPassengers") +
  theme_minimal()
```

**Reading the decomposition:**

- **Observed**: the raw series — clear upward trend and growing seasonal swings.
- **Trend**: the smoothed long-run growth, extracted by a moving average.
- **Seasonal**: the repeating annual pattern — a peak in summer, a trough in winter.
- **Random (residual)**: what remains after removing trend and seasonality. A well-specified model leaves mostly white noise here.

```{r ts_forecast, fig.height=5, fig.width=9}
# fit an ARIMA model automatically (auto.arima selects the best order)
# log transform: stabilises the variance (seasonal swings grow with level)
fit <- auto.arima(log(AirPassengers))

cat("Fitted model:\n")
print(fit)

# generate a 24-month (2-year) forecast with 80% and 95% prediction intervals
fc <- forecast(fit, h = 24)

# back-transform from log scale for the plot
autoplot(fc) +
  labs(
    title    = "24-month forecast for airline passengers (ARIMA on log scale)",
    subtitle = "Shaded bands = 80% and 95% prediction intervals",
    x        = "Year",
    y        = "log(Passengers)"
  ) +
  theme_minimal()
```

**Key points:** The model captures both trend (upward) and seasonality (annual cycle). The prediction intervals widen into the future — uncertainty grows the further ahead we forecast. This is an honest representation of what we can and cannot know from the data.

---

## 7. Social network analysis with R

### 7.1 What is social network analysis?

A **network** (or **graph**) consists of **nodes** (entities — people, genes, webpages) connected by **edges** (relationships — friendships, interactions, hyperlinks). Social network analysis (SNA) quantifies the structure of these relationships.

Key terms:

| Term | Definition |
|---|---|
| **Node / vertex** | An entity in the network |
| **Edge** | A relationship between two nodes. Can be directed (A → B) or undirected (A — B) |
| **Degree** | The number of edges connected to a node — a measure of local connectivity |
| **Betweenness centrality** | How often a node lies on the shortest path between other nodes. High betweenness = information broker. |
| **Closeness centrality** | The inverse of the average distance from a node to all others. High closeness = can reach others quickly. |
| **Transitivity** | The proportion of connected triples that form triangles (also called clustering coefficient) |
| **Community** | A subset of nodes that are more densely connected internally than to the rest of the network |

### 7.2 R packages for network analysis

| Package | Purpose |
|---|---|
| `igraph` | Core graph analysis: construction, centrality, clustering, visualisation |
| `sna` | Sociometric analysis; integrates with `network` objects |
| `ggraph` | `ggplot2`-based network visualisation |
| `tidygraph` | Tidy interface to graph data using `dplyr` verbs |

### 7.3 Example: the Zachary karate club network

Wayne Zachary's 1977 study is a canonical social network dataset. It records friendships between 34 members of a university karate club before the club split into two factions following a dispute between the instructor and the administrator. The network is built into `igraph`.

```{r sna_data}
# load the karate club network (built into igraph — no download needed)
karate <- igraph::make_graph("Zachary")

cat("Number of nodes (members):", vcount(karate), "\n")
cat("Number of edges (friendships):", ecount(karate), "\n")
cat("Network density:", round(edge_density(karate), 3),
    "(fraction of all possible edges that exist)\n")
cat("Average path length:", round(mean_distance(karate), 2), "\n")
cat("Transitivity (clustering coefficient):", round(transitivity(karate), 3), "\n")
```

```{r sna_centrality}
# compute centrality measures for each node
deg        <- degree(karate)            # degree centrality
between    <- betweenness(karate)       # betweenness centrality
close      <- closeness(karate)         # closeness centrality

# summarise top 5 nodes by betweenness (information brokers)
centrality_df <- tibble(
  node        = 1:vcount(karate),
  degree      = deg,
  betweenness = round(between, 1),
  closeness   = round(close, 4)
) %>%
  arrange(desc(betweenness))

cat("Top 5 nodes by betweenness centrality:\n")
print(head(centrality_df, 5))
```

```{r sna_communities}
# detect communities using the Walktrap algorithm
# (random walks tend to stay within communities)
communities <- cluster_walktrap(karate)

cat("Number of communities detected:", length(communities), "\n")
cat("Modularity score:", round(modularity(communities), 3),
    "(higher = more defined community structure)\n\n")
cat("Community membership:\n")
print(membership(communities))
```

```{r sna_plot, fig.height=7, fig.width=8}
# set node attributes for visualisation
V(karate)$community <- membership(communities)
V(karate)$degree    <- degree(karate)

# colour palette for communities
community_colours <- RColorBrewer::brewer.pal(
  max(V(karate)$community), "Set2"
)[V(karate)$community]

# plot the network
# node size ∝ degree; colour encodes community
set.seed(42)
plot(
  karate,
  vertex.color       = community_colours,
  vertex.size        = V(karate)$degree * 1.5,  # size ∝ degree
  vertex.label.cex   = 0.8,
  vertex.label.color = "black",
  edge.color         = "grey70",
  edge.width         = 0.8,
  layout             = layout_with_fr(karate),   # Fruchterman-Reingold layout
  main               = "Zachary karate club network\n(colour = community, size ∝ degree)"
)
legend("bottomleft",
       legend = paste("Community", 1:max(V(karate)$community)),
       fill   = RColorBrewer::brewer.pal(max(V(karate)$community), "Set2"),
       bty    = "n", cex = 0.8)
```

**What the network reveals:** Nodes 1 (instructor) and 34 (administrator) are the largest (highest degree) and sit at the boundary between the two main communities — they are the central figures around whom the club eventually split. Betweenness centrality identifies the nodes that serve as bridges between otherwise weakly connected parts of the network.

---

## 8. R and big data

### 8.1 When does "big data" become a problem?

Standard R loads all data into RAM. This works well for datasets up to a few hundred megabytes. Problems arise when:

- **Volume**: the dataset is too large to fit in memory (gigabytes or terabytes)
- **Velocity**: data arrives faster than a single machine can process it (streaming logs, sensor feeds)
- **Variety**: data comes in many formats — structured tables, JSON, images, text — often from multiple sources

These are the classic "3 Vs" of big data. Different tools address different Vs.

### 8.2 Current R big data ecosystem

| Tool | What it does | R interface |
|---|---|---|
| **Apache Spark** | Distributed computation across a cluster — handles Vs of volume and velocity | `sparklyr` (recommended), `SparkR` |
| **Apache Arrow** | In-memory columnar format; reads Parquet files and connects to cloud storage at high speed | `arrow` package |
| **DuckDB** | In-process analytical SQL database; handles files larger than RAM with minimal setup | `duckdb` package |
| **H2O** | In-memory distributed machine learning; scales to datasets too large for `caret` or `tidymodels` | `h2o` package |
| **MongoDB** | Document database (JSON-like records); good for unstructured or semi-structured data | `mongolite` package |

> **Note on deprecated packages:** Earlier versions of this course referenced `RHadoop`, `RHIPE`, `rmongodb`, and `RMongo`. These packages are no longer maintained and are not available on current CRAN. Use `sparklyr` for Hadoop/Spark integration and `mongolite` for MongoDB.

### 8.3 A practical decision guide

```
Is your data larger than available RAM?
├─ No  → Use tidyverse / tidymodels as normal.
└─ Yes → Does it fit on a single machine's disk?
         ├─ Yes → Use arrow (Parquet) or duckdb (SQL on files).
         └─ No  → Use sparklyr to distribute across a cluster.

Do you need machine learning at scale?
└─ Yes → Consider h2o (single machine, distributed RAM)
          or sparklyr with MLlib (cluster-scale).

Is your data in a document / NoSQL store?
└─ Yes → Use mongolite to read from MongoDB into R data frames.
```

### 8.4 A minimal `arrow` example

`arrow` is the easiest entry point for large-file work because it requires no server setup — it reads Parquet files directly from disk or cloud storage.

```{r bigdata_arrow, eval=FALSE}
# This chunk shows the pattern — not run because it requires a Parquet file.
# install.packages("arrow")
library(arrow)

# read a large Parquet file directly into R (lazy — only loads what you need)
ds <- open_dataset("path/to/large_file.parquet")

# use dplyr verbs before collecting into memory
result <- ds %>%
  filter(year == 2023) %>%
  group_by(region) %>%
  summarise(avg_sales = mean(sales, na.rm = TRUE)) %>%
  collect()   # <-- only now does data come into R memory
```

---

## 9. Exercises

---

### Exercise 1 — Extend the classification analysis

Using the iris classification example from Section 2:

**(a)** Fit a standard CART decision tree using `rpart::rpart()` with the same formula and training data. Plot the tree using `rpart.plot::rpart.plot()` (install if needed) and compare its structure to the `ctree` tree.

**(b)** Compute the test-set accuracy of the `rpart` tree. Which of the three classifiers — `ctree`, random forest, or `rpart` — performs best on the test set?

**(c)** Change `set.seed(465)` to `set.seed(99)` and re-run the split and all three classifiers. Do the rankings change? What does this tell you about evaluating classifiers on a single train/test split?

*Hint: For part (a), use `pred_rpart <- predict(rpart_fit, newdata = iris.test, type = "class")`. For part (c), consider how small test sets lead to high variance in accuracy estimates.*

<details>
<summary><strong>Solution — click to expand</strong></summary>

```{r ex1_solution, eval=TRUE}
library(rpart)

# Part (a): fit CART decision tree
rpart_fit <- rpart(
  Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
  data   = iris.train,
  method = "class"
)

# install.packages("rpart.plot")
# rpart.plot::rpart.plot(rpart_fit, main = "CART decision tree for iris species")

# Part (b): test-set accuracy for rpart
pred_rpart     <- predict(rpart_fit, newdata = iris.test, type = "class")
cm_rpart       <- table(Actual = iris.test$Species, Predicted = pred_rpart)
accuracy_rpart <- sum(diag(cm_rpart)) / sum(cm_rpart)

cat("--- Test-set accuracy comparison ---\n")
cat("rpart CART tree:             ", round(accuracy_rpart, 3), "\n")
cat("partykit ctree:              ", round(accuracy_ctree,  3), "\n")
cat("ranger random forest:        ", round(rf_accuracy,    3), "\n")
```

**Interpretation for part (a):** The `rpart` CART tree chooses splits using the Gini impurity criterion, which tends to prefer variables with many possible split points. The `ctree` uses statistical testing, which is more conservative — it may produce a simpler tree. Both should achieve similar accuracy on clean data like `iris`.

**Interpretation for part (b):** On `iris`, the random forest typically achieves the highest accuracy (often 95–100%) because averaging over 500 trees reduces the variance of individual splits. The two single-tree methods are more variable.

**Interpretation for part (c):** On a small test set (roughly 45 observations), a single misclassification changes accuracy by ~2 percentage points. Running with `set.seed(99)` will produce a different split, and the rankings may well change. This illustrates why k-fold cross-validation is preferred over a single split for small datasets — it averages performance across multiple held-out folds and gives a more stable estimate.

</details>

---

### Exercise 2 — Choosing K in k-means clustering

The k-means example in Section 3 used K = 3 because we knew the true number of species. In real data, we do not know K in advance.

**(a)** Apply the **elbow method**: compute WCSS (total within-cluster sum of squares) for K = 1 to 10 on the iris data (without the species column). Plot WCSS against K and identify the "elbow" — the point where adding more clusters gives diminishing returns.

**(b)** What value of K does the elbow suggest? Does it agree with the true number of species?

**(c)** Re-run k-means with your chosen K and compare the resulting clusters to the true species using a table. How does agreement change compared to K = 3?

*Hint: Use `sapply(1:10, function(k) kmeans(iris2, centers = k, nstart = 25)$tot.withinss)` to compute WCSS for each K.*

<details>
<summary><strong>Solution — click to expand</strong></summary>

```{r ex2_solution, eval=TRUE, fig.height=4, fig.width=6}
set.seed(42)
iris2_ex <- iris %>% dplyr::select(-Species)

# Part (a): compute WCSS for K = 1 to 10
wcss <- sapply(1:10, function(k) {
  kmeans(iris2_ex, centers = k, nstart = 25)$tot.withinss
})

# elbow plot
tibble(K = 1:10, WCSS = wcss) %>%
  ggplot(aes(x = K, y = WCSS)) +
  geom_line(colour = "steelblue") +
  geom_point(size = 3, colour = "steelblue") +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title    = "Elbow method: within-cluster SS vs. number of clusters (iris)",
    x        = "Number of clusters K",
    y        = "Total within-cluster SS"
  ) +
  theme_minimal()
```

**Part (b):** The elbow is typically visible at K = 3, which agrees with the true number of species. The curve drops steeply from K = 1 to K = 3, then flattens. This is a favourable case — `iris` has well-separated clusters. In real data the elbow is often less obvious, and analysts use multiple indices (silhouette score, gap statistic, NbClust) for confirmation.

```{r ex2_solution_k3, eval=TRUE}
# Part (c): K = 3 is the elbow — compare to true labels
set.seed(42)
km3 <- kmeans(iris2_ex, centers = 3, nstart = 25)
table(True_species = iris$Species, Cluster = km3$cluster)
```

With K = 3 and a different random seed, the cluster numbering may differ from the main example (cluster 1 in one run may correspond to cluster 3 in another — k-means labels are arbitrary). The pattern of which species overlap which clusters should be similar.

</details>

---

### Exercise 3 — Association rules: interpreting metrics

Using the rules mined from the `Groceries` dataset in Section 4:

**(a)** Find all rules where `whole milk` appears on the **right-hand side** (i.e., rules that predict whole milk purchases). How many such rules are there?

**(b)** Among those rules, which has the highest **confidence**? Write out the rule in plain English and explain what the confidence value means for a grocery store manager.

**(c)** A colleague claims "high lift always means a rule is useful for a business decision." Do you agree? Describe a scenario where a high-lift rule is not actionable.

*Hint: Use `subset(rules, rhs %in% "whole milk")` to filter rules by right-hand side item.*

<details>
<summary><strong>Solution — click to expand</strong></summary>

```{r ex3_solution, eval=TRUE}
# Part (a): rules with whole milk on the right-hand side
milk_rules <- subset(rules, rhs %in% "whole milk")
cat("Number of rules predicting 'whole milk':", length(milk_rules), "\n\n")

# Part (b): rule with highest confidence
cat("Top 5 whole-milk rules by confidence:\n")
arules::inspect(sort(milk_rules, by = "confidence")[1:5])
```

**Part (b) interpretation:** The rule with the highest confidence might read something like `{other vegetables, butter} → {whole milk}` with confidence 0.57. This means: of all transactions containing both other vegetables and butter, 57% also contained whole milk. For a store manager this suggests placing whole milk near the butter section, or creating a promotional bundle.

**Part (c):** High lift does not always mean actionable. Consider the rule `{baby food} → {diapers}` — lift may be very high (parents who buy baby food almost always buy diapers) but this rule adds little value to a manager because both items are already shelved together in the baby aisle and the association is entirely expected. Useful rules have high lift *and* high support *and* describe a non-obvious connection that suggests a concrete action (cross-sell, store layout change, promotion).

</details>

---

### Exercise 4 — Time series: decomposition and forecasting

The `nottem` dataset (built into R) contains monthly average temperatures at Nottingham Castle from 1920 to 1939.

**(a)** Plot the raw series. Describe the trend and seasonal pattern you observe.

**(b)** Decompose the series using `decompose(nottem, type = "additive")` and plot the result. Why is additive more appropriate here than multiplicative?

**(c)** Fit an ARIMA model using `auto.arima()` and produce a 36-month (3-year) forecast. Plot the forecast with prediction intervals. What is the model selected?

*Hint: Use `autoplot()` from the `forecast` package for all plots.*

<details>
<summary><strong>Solution — click to expand</strong></summary>

```{r ex4_solution, eval=TRUE, fig.height=4, fig.width=8}
data(nottem)

# Part (a): plot raw series
autoplot(nottem) +
  labs(title = "Monthly average temperature — Nottingham Castle (1920–1939)",
       x = "Year", y = "Temperature (°F)") +
  theme_minimal()
```

**Part (a):** The series shows strong seasonal variation (cold winters, warm summers) repeating every 12 months. There is no obvious long-run trend — the level appears roughly constant across the 20-year period.

```{r ex4_decompose, eval=TRUE, fig.height=6, fig.width=8}
# Part (b): additive decomposition
decomp_nottem <- stats::decompose(nottem, type = "additive")
autoplot(decomp_nottem) +
  labs(title = "Additive decomposition — Nottingham temperatures") +
  theme_minimal()
```

**Part (b):** Additive decomposition is appropriate because the **amplitude of the seasonal swing stays roughly constant** across the series — summer temperatures are always about 30°F warmer than winter temperatures, regardless of the overall level. A multiplicative model would be appropriate if the seasonal variation grew proportionally with the level (as in the airline passengers example, where traffic levels rose substantially over the period).

```{r ex4_forecast, eval=TRUE, fig.height=5, fig.width=8}
# Part (c): ARIMA model and forecast
fit_nottem <- auto.arima(nottem)
cat("Selected model:", capture.output(fit_nottem)[1], "\n")

fc_nottem <- forecast(fit_nottem, h = 36)
autoplot(fc_nottem) +
  labs(title    = "36-month forecast — Nottingham temperatures (ARIMA)",
       subtitle = "Shaded bands = 80% and 95% prediction intervals",
       x = "Year", y = "Temperature (°F)") +
  theme_minimal()
```

**Part (c):** `auto.arima()` typically selects a seasonal ARIMA model such as ARIMA(2,0,0)(1,1,0)[12], reflecting the strong annual seasonality (period = 12 months). Because the series is stationary in the mean, the forecast stays near the historical average with the familiar seasonal cycle. The prediction intervals widen over the 3-year horizon, reflecting growing uncertainty.

</details>

---

## Online references and further reading

### Statistical data mining — overview

- **R Data Mining** — tutorials, examples, and resources for data mining with R
  [http://www.rdatamining.com](http://www.rdatamining.com)

- **Introduction to Statistical Learning with R (2e)** — James, Witten, Hastie & Tibshirani *(free PDF)*
  [https://www.statlearning.com](https://www.statlearning.com)
  Chapters 8 (trees), 10 (unsupervised learning), and 12 (clustering) are directly relevant.

- **The Elements of Statistical Learning** — Hastie, Tibshirani & Friedman *(free PDF)*
  [https://hastie.su.domains/ElemStatLearn/](https://hastie.su.domains/ElemStatLearn/)
  The advanced companion to ISLR. Chapters 9 and 14 cover classification and association rules in depth.

### Classification and decision trees

- **`partykit` package vignette** — conditional inference trees and forests
  [https://cran.r-project.org/web/packages/partykit/vignettes/partykit.pdf](https://cran.r-project.org/web/packages/partykit/vignettes/partykit.pdf)

- **`ranger` package documentation** — fast random forests
  [https://cran.r-project.org/package=ranger](https://cran.r-project.org/package=ranger)

### Clustering

- **`cluster` package documentation**
  [https://cran.r-project.org/package=cluster](https://cran.r-project.org/package=cluster)

### Association rules

- **Introduction to `arules`** — Hahsler, Grün & Hornik
  [https://cran.r-project.org/web/packages/arules/vignettes/arules.pdf](https://cran.r-project.org/web/packages/arules/vignettes/arules.pdf)

### Text mining

- **Text Mining with R: A Tidy Approach** — Silge & Robinson *(free online)*
  [https://www.tidytextmining.com](https://www.tidytextmining.com)
  The definitive guide to the `tidytext` workflow. Covers tf-idf, topic modelling, and sentiment analysis.

### Time series

- **Forecasting: Principles and Practice (3e)** — Hyndman & Athanasopoulos *(free online)*
  [https://otexts.com/fpp3/](https://otexts.com/fpp3/)
  The authoritative reference for the `forecast` package and modern time series analysis in R.

### Social network analysis

- **`igraph` R package documentation**
  [https://igraph.org/r/](https://igraph.org/r/)

- **Network Science** — Barabási *(free online)*
  [http://networksciencebook.com](http://networksciencebook.com)
  Intuitive introduction to network theory with real-world examples.

### Big data with R

- **`sparklyr` documentation** — R interface to Apache Spark
  [https://spark.rstudio.com](https://spark.rstudio.com)

- **`arrow` R package documentation**
  [https://arrow.apache.org/docs/r/](https://arrow.apache.org/docs/r/)

- **`duckdb` R package documentation**
  [https://duckdb.org/docs/api/r.html](https://duckdb.org/docs/api/r.html)

---

## Glossary

**Accuracy:** The proportion of all test observations that were classified correctly: (TP + TN) / Total.

**ARIMA:** AutoRegressive Integrated Moving Average. A class of time series models that combine autoregression (current value depends on past values), differencing (to achieve stationarity), and moving averages (current value depends on past forecast errors).

**Association rule:** An if-then statement of the form {A} → {B} expressing that items A and B co-occur frequently in transactions.

**Bag of words:** A document representation that records only word counts, ignoring word order. Forms the basis of many text classification and clustering methods.

**Betweenness centrality:** A node-level measure of how often a node lies on the shortest path between other node pairs. High betweenness = an information broker or bridge in the network.

**Centroid:** The mean position of all observations assigned to a cluster — the cluster's centre of gravity. k-means relocates centroids iteratively until assignments stabilise.

**Closeness centrality:** The inverse of the average shortest path distance from a node to all other nodes. High closeness = can spread information quickly to the whole network.

**Community (network):** A subset of nodes with denser internal connections than connections to the rest of the graph — a functional subgroup within a network.

**Confidence (association rules):** The proportion of transactions containing the left-hand side (LHS) that also contain the right-hand side (RHS). Measures the reliability of a rule.

**Confusion matrix:** A table comparing actual class labels to predicted labels, showing true positives, true negatives, false positives, and false negatives.

**Corpus:** In text mining, the full collection of text documents being analysed.

**Decision tree:** A classification or regression model that makes predictions by applying a sequence of binary split rules, forming a tree structure.

**Decomposition (time series):** Separating a time series into trend, seasonal, and residual components, each of which can then be analysed or modelled separately.

**Degree (network):** The number of edges connected to a node. In a directed network, in-degree and out-degree are counted separately.

**Edge:** A connection between two nodes in a network. Can be directed (with a source and target) or undirected.

**k-means:** A clustering algorithm that partitions observations into K groups by iteratively assigning each observation to its nearest centroid and then recomputing centroids.

**Lift (association rules):** The ratio of the observed co-occurrence frequency of two items to what would be expected if they were statistically independent. Lift > 1 indicates a genuine positive association.

**Node (vertex):** A fundamental unit of a network — an entity such as a person, webpage, or gene.

**Random forest:** An ensemble learning method that fits many decision trees, each on a random bootstrap sample with a random feature subset, and combines predictions by majority vote (classification) or averaging (regression).

**Seasonality:** A repeating pattern in a time series at a fixed, known period (e.g., annual, quarterly, monthly).

**Stationarity:** A property of a time series whose statistical characteristics — mean, variance, autocorrelation — do not change over time. Required by many time series models.

**Stop words:** High-frequency words ("the", "is", "and") that carry little semantic meaning and are typically removed early in a text mining pipeline.

**Support (association rules):** The proportion of all transactions that contain a specific itemset. Measures how common an itemset or rule is.

**TF-IDF:** Term Frequency–Inverse Document Frequency. A word-weighting scheme that gives high weight to words that are frequent in one document but rare across the corpus — identifying a document's distinctive vocabulary.

**Tokenisation:** Splitting a text document into individual units (tokens), typically words, for further analysis.

**Transitivity (clustering coefficient):** In a network, the proportion of connected triples that close into triangles. High transitivity suggests the network has dense local groupings.

**WCSS (within-cluster sum of squares):** The total squared distance of every observation from the centroid of its assigned cluster. k-means minimises WCSS. Also called inertia.
