---
title: 'Week 13 - Machine Learning: Supervised Learning'
date: "`r Sys.Date()`"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*(Notes adapted from **Modern Data Science with R** book)*

------------------------------------------------------------------------

## Learning objectives

By the end of this notebook, you will be able to:

> Build and evaluate a **k-nearest neighbors (k-NN)** classifier in R using `tidymodels`, and explain how the choice of `k` affects model accuracy.

> Apply a **Naive Bayes** classifier to a real dataset, interpret the resulting confusion matrix, and calculate predicted probabilities by hand using Bayes' theorem.

> Fit a **logistic regression** model for a binary outcome, interpret the model coefficients as odds ratios, and assess model performance using a cutoff rule.

------------------------------------------------------------------------

## Packages used in this notebook

```{r packages, warning = FALSE, message = FALSE}
#install.packages("pkgbuild")
#pkgbuild::check_rtools(debug = TRUE)

library(mdsr)       # datasets and utilities from Modern Data Science with R
library(ggplot2)    # data visualization
library(tidyverse)  # data wrangling (dplyr, tidyr, etc.)
#install.packages("tidymodels", dependencies = TRUE)
library(tidymodels) # unified interface for model fitting and evaluation
library(discrim)    # Naive Bayes classifier via naive_Bayes()
library(klaR)       # engine used by discrim for Naive Bayes
library(kknn)       # engine used for k-NN classification
```

------------------------------------------------------------------------

## What is supervised classification?

Before we dive in, let's set the stage. **Supervised learning** means we train a model using data that already has correct labels — in our case, whether a person earns more or less than \$50,000. The model learns patterns from this *training* data and then makes predictions on new, unseen *test* data.

A **classifier** is a model that predicts which *category* (or *class*) a new observation belongs to. In this notebook, all three methods — k-NN, Naive Bayes, and logistic regression — are classifiers that predict a binary outcome (e.g., income above or below \$50k, or diabetic vs. not diabetic).

A key principle in machine learning is never to evaluate your model on the same data you used to train it — that would be like studying the exact exam questions beforehand. Instead, we always hold out a **test set** to get an honest measure of performance.

------------------------------------------------------------------------

## Motivation: high-earners in the 1994 United States Census

A marketing analyst might be interested in finding factors that can be used to predict whether a potential customer is a high-earner. The 1994 United States Census provides information that can inform such a model, with records from 32,561 adults that include a binary variable indicating whether each person makes greater or less than \$50,000 (more than \$80,000 today after accounting for inflation). This is our **response variable**.

### Data preparation

```{r prepare data, message=FALSE}
library(tidyverse)
library(mdsr)

url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data"

census <- read_csv(
  url,
  col_names = c(
    "age", "workclass", "fnlwgt", "education",
    "education_1", "marital_status", "occupation", "relationship",
    "race", "sex", "capital_gain", "capital_loss", "hours_per_week",
    "native_country", "income"
  )
) %>%
  mutate(income = factor(income))  # convert income to a factor (categorical variable)

# inspect the structure: variable names, types, and first few values
glimpse(census)
```

### Train/test split

For reasons we will discuss shortly, we separate the data into two pieces at random: 80% becomes the **training** set (used to build the model) and 20% becomes the **test** set (used to evaluate the model on unseen data).

```{r split data set}
library(tidymodels)
set.seed(364)          # set seed for reproducibility — ensures the same split each run

n <- nrow(census)
census_parts <- census %>%
  initial_split(prop = 0.8)       # 80/20 split

train <- census_parts %>% training()   # extract training data
test  <- census_parts %>% testing()    # extract test (hold-out) data

# calculate the proportion of high earners in the training set
pi_bar <- train %>%
  count(income) %>%
  mutate(pct = n / sum(n)) %>%
  filter(income == ">50K") %>%
  pull(pct)

print(c("Percent earning >$50K in training set:", round(pi_bar, 3)))
```

Note that only about 24% of those in the sample make more than \$50k. This means the **null model** — a model that simply predicts "≤\$50K" for everyone — would be correct about **76%** of the time. This is our baseline to beat.

**Pro tip:** Always benchmark your predictive models against a reasonable null model. An accuracy of 77% sounds good until you realise a naive guess achieves 76%.

### Model formula

We define the set of predictor variables using a formula object. This formula is reused across all three classifiers below.

```{r model form}
form <- as.formula(
  "income ~ age + workclass + education + marital_status +
  occupation + relationship + race + sex +
  capital_gain + capital_loss + hours_per_week"
)

form
```

------------------------------------------------------------------------

## k-Nearest Neighbors (k-NN) classifier

### What is k-NN?

The **k-nearest neighbors** algorithm is a non-parametric, supervised learning classifier that makes predictions based on proximity. To classify a new observation, k-NN finds the `k` most similar observations in the training data (its "neighbors") and assigns the most common label among them.

Key properties:

-   Can be used for **regression or classification**; most commonly used for classification.
-   Requires no model training time — the entire dataset is stored and processed at prediction time (so predictions can be slow on large datasets).
-   Requires a **distance metric**. Euclidean distance is the most common choice.
-   Only works with **numeric variables** (distance is undefined for categories without encoding).
-   The choice of `k` matters: small `k` = flexible but noisy; large `k` = smoother but may underfit.

### Algorithm for classification

1.  For a new observation, compute its distance to every point in the training set.
2.  Identify the `k` closest neighbors.
3.  Assign the label that appears most often among those `k` neighbors (majority vote).

**Note on "majority vote":** Technically this is *plurality* voting — the most frequent label wins, even if it has fewer than 50% of the `k` votes. This distinction matters when there are more than two classes.

### Example: applying k-NN to the Census data

Since Euclidean distance only works with numeric variables, we first select only the numeric predictors.

```{r knn-fit}
library(kknn)

# select numeric predictors only (distance metrics don't work on character/factor variables)
# fnlwgt is a survey weight, not a meaningful predictor, so we exclude it
train_q <- train %>%
  dplyr::select(income, where(is.numeric), -fnlwgt)

# fit a k-NN classifier with k=5
# scale=TRUE standardizes each variable to have mean 0 and SD 1
# this prevents variables with large scales (e.g. capital_gain) from dominating distance
mod_knn <- nearest_neighbor(neighbors = 5, mode = "classification") %>%
  set_engine("kknn", scale = TRUE) %>%
  fit(income ~ ., data = train_q)

# generate predictions on the training set and attach them to the data
pred <- train_q %>%
  bind_cols(
    predict(mod_knn, new_data = train_q, type = "class")
  ) %>%
  rename(income_knn = .pred_class)   # rename the predicted column for clarity

# display the confusion matrix
pred %>%
  conf_mat(income, income_knn)
```

#### Reading the confusion matrix

The confusion matrix compares actual labels (rows) to predicted labels (columns):

|                    | Predicted ≤\$50K    | Predicted \>\$50K   |
|--------------------|---------------------|---------------------|
| **Actual ≤\$50K**  | True Negative (TN)  | False Positive (FP) |
| **Actual \>\$50K** | False Negative (FN) | True Positive (TP)  |

-   **True Positives (TP):** Correctly predicted high earners.
-   **True Negatives (TN):** Correctly predicted non-high earners.
-   **False Positives (FP):** Predicted high earner, but actually earned ≤\$50K (Type I error).
-   **False Negatives (FN):** Predicted ≤\$50K, but actually earned \>\$50K (Type II error).

**Accuracy** = (TP + TN) / Total — the proportion of all predictions that were correct.

```{r knn-accuracy}
# accuracy = (true positives + true negatives) / total observations
pred %>%
  accuracy(income, income_knn)
```

*Expected result: accuracy around 83–85% on the training set.*

### Choosing the right `k`

The choice of `k` affects model performance significantly. We can evaluate different values of `k` using a process called **cross-validation** — fitting the model at each `k` and measuring accuracy on the training data.

```{r knn-tune, warning=FALSE}
# helper function: fit a k-NN model for a given k
knn_fit <- function(.data, k) {
  nearest_neighbor(neighbors = k, mode = "classification") %>%
    set_engine("kknn", scale = TRUE) %>%
    fit(income ~ ., data = .data)
}

# helper function: evaluate accuracy of a fitted model on new data
knn_accuracy <- function(mod, .new_data) {
  mod %>%
    predict(new_data = .new_data) %>%
    mutate(income = .new_data$income) %>%
    accuracy(income, .pred_class) %>%
    pull(.estimate)
}

# values of k to evaluate
ks <- c(1:10, 15, 20, 30, 40, 50)

# NOTE: this block is computationally intensive.
# Comment it out after running once to speed up re-knitting.
knn_tune <- tibble(
  k = ks,
  mod = map(k, knn_fit, .data = train_q),                     # fit a model at each k
  train_accuracy = map_dbl(mod, knn_accuracy, .new_data = train_q)  # evaluate each model
)

knn_tune
```

```{r knn-plot}
# plot accuracy vs k — look for the "elbow" where accuracy stops improving
ggplot(data = knn_tune, aes(x = k, y = train_accuracy)) +
  geom_point() +
  geom_line() +
  labs(
    title = "k-NN training accuracy by number of neighbors",
    x = "k (number of neighbors)",
    y = "Accuracy"
  )
```

**Interpreting the plot:** Accuracy tends to decrease as `k` increases on this dataset, suggesting smaller values of `k` fit the training data better. However, be cautious: a very small `k` (especially `k=1`) may **overfit** — it memorizes the training data but may generalize poorly to new data. On the test set, a moderate `k` often performs best.

**Think about it:** Why might `k=1` achieve near-perfect training accuracy but perform worse on the test set than `k=5`?

------------------------------------------------------------------------

## Naive Bayes classifier

### What is Bayes' theorem?

The Naive Bayes classifier is grounded in **Bayes' theorem**, a fundamental result in probability:

$$P(y \mid x) = \frac{P(x \mid y) \cdot P(y)}{P(x)}$$

In plain English: *The probability of a label* $y$ given our observed features $x$ equals how often those features appear alongside that label, scaled by how common the label is overall, divided by how often those features appear in general.

Each term has a name:

| Term | Name | Meaning |
|----|----|----|
| $P(y \mid x)$ | Posterior | Probability of the label **given** what we observed |
| $P(x \mid y)$ | Likelihood | Probability of observing these features **given** the label |
| $P(y)$ | Prior | Overall probability of the label (before seeing any features) |
| $P(x)$ | Evidence | Overall probability of observing these features |

### How Naive Bayes classifies

To classify a new observation $x^*$, we compare:

$$P(y = 1 \mid x^*) \quad \text{vs} \quad P(y = 0 \mid x^*)$$

If $P(y = 1 \mid x^*) > P(y = 0 \mid x^*)$, we predict the label $y = 1$.

The "naive" assumption is that all predictor variables are **conditionally independent** given the label. This is rarely true in practice (e.g., education and occupation are correlated), but the classifier often performs surprisingly well despite this simplification — hence the name *naive*.

### Worked example: Bayes' theorem by hand

Consider the first person in the training data — a 39-year-old white male with a bachelor's degree working in a clerical role for state government. In reality, this person earned ≤\$50K.

```{r nb-head}
# look at the first row of the training data
train %>%
  as.data.frame() %>%
  head(1)
```

Using our training data, we can estimate:

-   $P(>50K) = 0.243$ — about 24% of people in the training set earn more than \$50K.
-   $P(\text{male} \mid >50K) = 0.845$ — among high earners, 84.5% are male.
-   $P(\text{male}) = 0.670$ — overall, 67% of people in the dataset are male.

Applying Bayes' theorem:

$$P(>50K \mid \text{male}) = \frac{P(\text{male} \mid >50K) \cdot P(>50K)}{P(\text{male})} = \frac{0.845 \times 0.243}{0.670} = 0.306$$

So, knowing a person is male raises the predicted probability of high income from 24.3% to about 30.6%.

```{r nb-verify}
# verify P(male) — proportion of males in the training set
t1 <- train %>% pull(sex) %>% table()
prop.table(t1)

# verify P(male | >50K) — proportion of males among high earners
t2 <- train %>% filter(income == ">50K") %>% pull(sex) %>% table()
prop.table(t2)
```

In practice, Naive Bayes extends this to many variables simultaneously, multiplying the conditional probabilities for each feature (under the independence assumption).

### Fitting the Naive Bayes classifier

```{r nb-fit, warning=FALSE}
library(discrim)

# fit a Naive Bayes classifier using the full formula (includes categorical predictors)
mod_nb <- naive_Bayes(mode = "classification") %>%
  set_engine("klaR") %>%        # klaR is the underlying engine
  fit(form, data = train)       # train on the training set

# generate predictions on the training set
pred <- train %>%
  bind_cols(
    predict(mod_nb, new_data = train, type = "class")
  ) %>%
  rename(income_nb = .pred_class)

# display the confusion matrix
pred %>%
  conf_mat(income, income_nb)

# calculate and display accuracy
pred %>%
  accuracy(income, income_nb)
```

*Expected result: accuracy around 82–84% on the training set.*

**Exercise:** Apply the trained Naive Bayes classifier (`mod_nb`) to the `test` set and compare the accuracy. Does it perform better or worse than on the training set? Why might this be?

*Hint: use `predict(mod_nb, new_data = test, type = "class")` and then join and evaluate as above.*

**Think about it:** The Naive Bayes model uses categorical variables (like `sex`, `occupation`, `race`) that the k-NN model had to drop. How might this advantage affect its predictions?

------------------------------------------------------------------------

## Logistic regression classifier

### Key terms

| Term | Definition |
|----|----|
| **Odds** | The ratio of the probability of success to the probability of failure: $\frac{p}{1-p}$ |
| **Log odds (logit)** | The natural log of the odds: $\log\!\left(\frac{p}{1-p}\right)$. Maps probabilities from $(0,1)$ to $(-\infty, +\infty)$ |
| **Logistic response function** | The inverse of the logit; maps any real number back to a probability in $(0, 1)$ |
| **Odds ratio** | The ratio of two odds values; used to interpret the effect of a predictor on the outcome |

### The logistic response function and the logit

With binary outcomes, we cannot apply ordinary linear regression directly — predicted values could fall outside the $[0,1]$ range required for probabilities. Instead, we model the **probability** $p$ that the label is 1 using the **logistic response function**:

$$p = \frac{e^{\beta_0 + \beta_1 x_1 + \cdots + \beta_q x_q}}{1 + e^{\beta_0 + \beta_1 x_1 + \cdots + \beta_q x_q}}$$

**Q:** What are the limits of $p$ (upper and lower) given this form? *(Answer:* $p$ is always between 0 and 1, no matter how large or small the linear combination gets.)

To work with a linear model, we rearrange using **odds**. If the probability of success is $p$, the odds are:

$$\text{Odds}(Y=1) = \frac{p}{1-p}$$

For example: if the probability of winning is $p = 0.5$, the odds are $0.5 / 0.5 = 1.0$ (even odds). If $p = 0.75$, the odds are $0.75 / 0.25 = 3.0$ (three-to-one).

Taking the natural log of the odds gives us the **logit** (or log-odds), which is linear in the predictors:

$$\log\!\left(\text{Odds}(Y=1)\right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_q x_q$$

This is the linear model we estimate. After fitting, we can convert predicted log-odds back to probabilities using $p = \frac{\text{Odds}}{1 + \text{Odds}}$. Finally, we classify each observation by applying a **cutoff rule**: if the predicted probability exceeds some threshold (commonly 0.5), predict class 1.

### Example: predicting diabetes from age and BMI

We now use the `NHANES` dataset to model the probability of having diabetes as a function of `Age` and `BMI`.

```{r logistic-plot, warning=FALSE}
library(NHANES)
data(NHANES)

# create a binary numeric outcome (1 = has diabetes, 0 = does not)
NHANES <- NHANES %>%
  mutate(has_diabetes = as.numeric(Diabetes == "Yes"))

# plot diabetes status vs Age with a logistic regression smoother
log_plot <- ggplot(data = NHANES, aes(x = Age, y = has_diabetes)) +
  geom_jitter(alpha = 0.1, height = 0.05) +               # jitter to reduce overplotting
  geom_smooth(method = "glm",
              method.args = list(family = "binomial")) +   # fit a logistic curve
  ylab("Diabetes status (1 = Yes, 0 = No)")

log_plot + xlab("Age (in years)")

log_plot +
  aes(x = BMI) +
  xlab("BMI (body mass index)")
```

**Q:** Which variable appears more strongly associated with diabetes: `Age` or `BMI`?

### Fitting the logistic regression model

We use `glm()` with `family = "binomial"` to fit a logistic regression model. The `binomial` family tells R that the response is binary and to use the logistic link function.

```{r logistic-fit}
# fit logistic regression with both predictors
logreg <- glm(has_diabetes ~ BMI + Age, family = "binomial", data = NHANES)

# display model coefficients with standard errors and p-values
tidy(logreg)
```

Both `BMI` and `Age` are statistically significant predictors (p \< 0.05), meaning we have strong evidence that each is associated with diabetes risk.

### Predicted probabilities

The model produces **log-odds** (logit) values by default. We convert them to probabilities manually:

```{r logistic-probs}
# predict() returns log-odds (the linear component) by default
pred <- predict(logreg)

# convert log-odds to probabilities using the inverse logit formula: p = 1 / (1 + exp(-logit))
prob <- 1 / (1 + exp(-pred))

# inspect a few predicted probabilities
prob[100:107]

# summary of the predicted probability distribution
summary(prob)
```

### Visualizing predicted probabilities across Age and BMI

```{r logistic-grid, warning=FALSE}
library(modelr)

# create a grid of all Age × BMI combinations for smooth visualization
fake_grid <- data_grid(
  NHANES,
  Age = seq_range(Age, 100),   # 100 evenly spaced values across the Age range
  BMI = seq_range(BMI, 100)    # 100 evenly spaced values across the BMI range
)

# get predicted probabilities for every point on the grid
y_hats <- fake_grid %>%
  mutate(y_hat = predict(logreg, newdata = ., type = "response"))  # type="response" gives probabilities directly

head(y_hats, 1)
```

```{r logistic-heatmap}
# visualize predicted probability surface: red = higher diabetes probability
ggplot(data = NHANES, aes(x = Age, y = BMI)) +
  geom_tile(data = y_hats, aes(fill = y_hat), color = NA) +
  geom_count(aes(color = factor(has_diabetes)), alpha = 0.8) +
  scale_fill_gradient(low = "white", high = "red") +
  scale_color_manual("Diabetes", values = c("gold", "black")) +
  scale_size(range = c(0, 2)) +
  labs(
    fill = "Predicted\nprobability",
    title = "Predicted probability of diabetes by Age and BMI"
  )
```

The red region (upper right) represents older individuals with higher BMI — the group with the highest predicted probability of diabetes.

### Manual probability calculation

For a hypothetical subject aged 0 with BMI 12.9 (the first point in our grid):

```{r logistic-manual-young}
# compute the linear component (log-odds) using matrix multiplication
linear_component <- c(1, 12.9, 0) %*% coef(logreg)  # c(intercept, BMI, Age) · coefficients

# convert log-odds to probability
exp(linear_component) / (1 + exp(linear_component))
```

Very few young people have diabetes, even with moderately elevated BMI. The predicted probability is about 0.1%.

What about a 60-year-old with a BMI of 25?

```{r logistic-manual-older}
linear_component <- c(1, 25, 60) %*% coef(logreg)
exp(linear_component) / (1 + exp(linear_component))
```

The predicted probability is meaningfully higher — illustrating how both age and BMI combine to increase risk.

### Interpreting odds ratios

One key advantage of logistic regression is **interpretability**. The coefficient $\beta_i$ for predictor $X_i$ represents the log of the **odds ratio** — the multiplicative change in the odds of $Y = 1$ for a one-unit increase in $X_i$, holding all other variables constant.

For a binary predictor $X$ (0 or 1), the odds ratio is:

$$\text{Odds ratio} = \frac{\text{Odds}(Y=1 \mid X=1)}{\text{Odds}(Y=1 \mid X=0)}$$

An odds ratio of 2 means the outcome is twice as likely when $X = 1$ versus $X = 0$.

For a numeric predictor like `BMI`, the odds ratio reflects a **one-unit increase**. The `BMI` coefficient from the model is approximately 0.0943, so:

```{r logistic-or}
# exponentiate the BMI coefficient to get the odds ratio
odds_ratio_BMI <- exp(0.094325)
odds_ratio_BMI
```

A one-unit increase in BMI multiplies the odds of diabetes by approximately **1.099** — about a 10% increase in odds per BMI unit, holding age constant.

### Assessing logistic regression performance

We use a **cutoff of 0.5**: if the predicted probability exceeds 0.5, we predict diabetes; otherwise, we predict no diabetes.

```{r logistic-accuracy}
# get predicted probabilities for the full NHANES dataset
logit_pred_prob <- predict(logreg, newdata = NHANES, type = "response")

# classify observations: 1 if predicted probability > 0.5, else 0
pred_y <- as.numeric(logit_pred_prob > 0.5)

# confusion matrix
confusion <- table(pred_y, NHANES$has_diabetes)
confusion

# overall accuracy: proportion of correct predictions
mean(pred_y == NHANES$has_diabetes, na.rm = TRUE)
```

*Expected result: accuracy around 91–92%. Note: the high accuracy is partly driven by the fact that most people do not have diabetes (the null model already achieves \~89%). Inspect the confusion matrix carefully — how well does the model detect actual diabetic cases?*

**Note on the cutoff:** The threshold of 0.5 is a common default, but it is not always optimal. A lower cutoff (e.g., 0.3) would classify more people as diabetic — catching more true cases (higher sensitivity) but also making more false alarms (lower specificity). The right cutoff depends on the cost of each type of error in your application.

**Think about it:** In a medical context, which error is more costly — a false positive (predicting diabetes when the patient is healthy) or a false negative (missing a true diabetic case)? How would you adjust the cutoff accordingly?

------------------------------------------------------------------------

## Exercises

------------------------------------------------------------------------

### Exercise 1 — Naive Bayes on the test set

Apply the trained Naive Bayes classifier (`mod_nb`) to the `test` set. Report the confusion matrix and accuracy.

*Hint: use `predict(mod_nb, new_data = test, type = "class")`, then join to the test data and call `conf_mat()` and `accuracy()`.*

**Think about it:** Does accuracy go up or down compared to the training set result? Why would you expect a difference?

<details>

<summary><strong>Solution — click to expand</strong></summary>

```{r ex1-solution, warning=FALSE}
# apply the trained Naive Bayes model to the held-out test set
pred_nb_test <- test %>%
  bind_cols(
    predict(mod_nb, new_data = test, type = "class")
  ) %>%
  rename(income_nb = .pred_class)   # rename predicted column for clarity

# confusion matrix on the test set
pred_nb_test %>%
  conf_mat(income, income_nb)

# accuracy on the test set
pred_nb_test %>%
  accuracy(income, income_nb)
```

**Interpretation:** You should see test-set accuracy in the range of **81–83%**, slightly lower than the training accuracy (\~83–84%). This is expected: models almost always perform a little worse on data they have never seen. The drop is small here, which suggests the Naive Bayes classifier generalises reasonably well and is not badly overfitting.

Looking at the confusion matrix, notice that the model tends to have more difficulty correctly identifying high earners (\>\$50K) than low earners — this is typical when the positive class is the minority (only \~24% of the data). A model that struggles to detect the minority class may still report high overall accuracy because it gets the majority class right most of the time.

</details>

------------------------------------------------------------------------

### Exercise 2 — Logistic regression on a NHANES train/test split

Split the original `NHANES` data into training (80%) and test (20%) sets. Fit a logistic regression model predicting `has_diabetes` from `BMI` and `Age` using only the training set. Report the accuracy on the test set.

*Hint: use `initial_split(NHANES, prop = 0.8)` followed by `training()` and `testing()`.*

**Think about it:** How does the test-set accuracy compare to the in-sample accuracy reported earlier (\~91–92%)? What does any difference tell you about the model?

<details>

<summary><strong>Solution — click to expand</strong></summary>

```{r ex2-solution, warning=FALSE}
# ensure NHANES has the binary diabetes variable (created earlier in the notebook)
NHANES <- NHANES %>%
  mutate(has_diabetes = as.numeric(Diabetes == "Yes"))

# set seed for reproducibility
set.seed(123)

# split NHANES into 80% training and 20% test
nhanes_split <- initial_split(NHANES, prop = 0.8)
nhanes_train <- training(nhanes_split)
nhanes_test  <- testing(nhanes_split)

# fit logistic regression on the training set only
logreg_train <- glm(
  has_diabetes ~ BMI + Age,
  family  = "binomial",
  data    = nhanes_train,
  na.action = na.omit   # drop rows with missing BMI or Age
)

# display the model coefficients
tidy(logreg_train)

# predict on the test set (type="response" gives probabilities directly)
test_probs <- predict(logreg_train, newdata = nhanes_test, type = "response")

# classify using a 0.5 cutoff
test_pred_y <- as.numeric(test_probs > 0.5)

# confusion matrix on the test set
confusion_test <- table(
  predicted = test_pred_y,
  actual    = nhanes_test$has_diabetes
)
confusion_test

# test-set accuracy
mean(test_pred_y == nhanes_test$has_diabetes, na.rm = TRUE)
```

**Interpretation:** Test-set accuracy should be around **91–92%**, very close to the in-sample figure. This tells us the logistic regression model is stable — it is not memorising the training data and generalises well to new observations.

However, inspect the confusion matrix closely. The model correctly identifies almost all non-diabetic cases (the majority class) but misses a substantial proportion of actual diabetic cases (false negatives). High overall accuracy can be misleading when one class dominates. In a clinical setting, missing a diabetic diagnosis is a serious error, so a lower cutoff (e.g., 0.3) might be preferred even though it would reduce overall accuracy.

</details>

------------------------------------------------------------------------

### Exercise 3 — Comparing all three classifiers on NHANES

Using the same training and test split from Exercise 2, fit k-NN and Naive Bayes classifiers. Compare the test-set accuracy of all three methods (logistic regression, k-NN, Naive Bayes) in a summary table.

*Hint: for k-NN, select only numeric variables. For Naive Bayes, convert `has_diabetes` to a factor before fitting.*

**Think about it:** Which method performs best? Does the ranking change if you focus on false negatives instead of overall accuracy?

<details>

<summary><strong>Solution — click to expand</strong></summary>

```{r ex3-solution, warning=FALSE, message=FALSE}
# ── Step 1: prepare data ──────────────────────────────────────────────────────

# for Naive Bayes and logistic regression, has_diabetes must be a factor
nhanes_train_f <- nhanes_train %>%
  mutate(has_diabetes = factor(has_diabetes))   # convert to factor for classification

nhanes_test_f <- nhanes_test %>%
  mutate(has_diabetes = factor(has_diabetes))

# for k-NN, keep only numeric predictors (distance metrics require numeric input)
nhanes_train_q <- nhanes_train %>%
  dplyr::select(has_diabetes, BMI, Age) %>%
  mutate(has_diabetes = factor(has_diabetes)) %>%
  drop_na()   # k-NN cannot handle missing values

nhanes_test_q <- nhanes_test %>%
  dplyr::select(has_diabetes, BMI, Age) %>%
  mutate(has_diabetes = factor(has_diabetes)) %>%
  drop_na()

# ── Step 2: fit k-NN (k=5) ───────────────────────────────────────────────────

mod_knn_nhanes <- nearest_neighbor(neighbors = 5, mode = "classification") %>%
  set_engine("kknn", scale = TRUE) %>%
  fit(has_diabetes ~ BMI + Age, data = nhanes_train_q)

# predict on the test set
knn_acc <- nhanes_test_q %>%
  bind_cols(
    predict(mod_knn_nhanes, new_data = nhanes_test_q, type = "class")
  ) %>%
  rename(pred = .pred_class) %>%
  accuracy(has_diabetes, pred) %>%
  pull(.estimate)

# confusion matrix for k-NN
nhanes_test_q %>%
  bind_cols(predict(mod_knn_nhanes, new_data = nhanes_test_q, type = "class")) %>%
  rename(pred = .pred_class) %>%
  conf_mat(has_diabetes, pred)

# ── Step 3: fit Naive Bayes ───────────────────────────────────────────────────

mod_nb_nhanes <- naive_Bayes(mode = "classification") %>%
  set_engine("klaR") %>%
  fit(has_diabetes ~ BMI + Age, data = nhanes_train_f %>% drop_na(BMI, Age))

# predict on the test set
nb_acc <- nhanes_test_f %>%
  drop_na(BMI, Age) %>%
  bind_cols(
    predict(mod_nb_nhanes,
            new_data = nhanes_test_f %>% drop_na(BMI, Age),
            type = "class")
  ) %>%
  rename(pred = .pred_class) %>%
  accuracy(has_diabetes, pred) %>%
  pull(.estimate)

# confusion matrix for Naive Bayes
nhanes_test_f %>%
  drop_na(BMI, Age) %>%
  bind_cols(predict(mod_nb_nhanes,
                    new_data = nhanes_test_f %>% drop_na(BMI, Age),
                    type = "class")) %>%
  rename(pred = .pred_class) %>%
  conf_mat(has_diabetes, pred)

# ── Step 4: logistic regression accuracy (from Exercise 2) ───────────────────

logreg_acc <- mean(test_pred_y == nhanes_test$has_diabetes, na.rm = TRUE)

# ── Step 5: summary comparison table ─────────────────────────────────────────

comparison <- tibble(
  Method   = c("k-NN (k=5)", "Naive Bayes", "Logistic Regression"),
  Test_Accuracy = round(c(knn_acc, nb_acc, logreg_acc), 4)
)

comparison
```

**Interpretation:** All three classifiers achieve similar overall accuracy (\~89–92%) on this dataset, which reflects the fact that the majority class (no diabetes, \~89% of records) dominates the accuracy metric.

Where the methods differ more meaningfully is in how they handle the minority class (diabetic cases):

-   **Logistic regression** typically produces well-calibrated probabilities and is the most interpretable — you can quantify how much each year of age or BMI unit increases diabetes risk.
-   **Naive Bayes** can perform competitively despite its simplifying independence assumption, and is very fast to fit.
-   **k-NN** with only two numeric predictors (BMI and Age) has less information than the other two models on this dataset, which may hurt its ability to detect diabetics.

The "best" method depends on your goal. If overall accuracy is all that matters, all three are similar. If detecting diabetic cases (minimising false negatives) matters more — as it would in a clinical setting — you should compare **sensitivity** (true positive rate) across the three methods, not just accuracy.

</details>

------------------------------------------------------------------------

## Comparison of the three classifiers

| Feature | k-NN | Naive Bayes | Logistic Regression |
|----|----|----|----|
| **Numeric predictors** | Required | Optional | Optional |
| **Categorical predictors** | Not directly | Yes | Yes (as dummies) |
| **Training time** | None (lazy learner) | Fast | Fast |
| **Prediction time** | Slow (scans all data) | Fast | Fast |
| **Interpretable coefficients** | No | Partial | Yes (odds ratios) |
| **Sensitive to variable scale** | Yes — use `scale=TRUE` | No | Somewhat |
| **Key tuning parameter** | `k` (number of neighbors) | Smoothing | Cutoff threshold |
| **Works best when...** | Decision boundary is complex and local | Features are roughly independent | Relationship is approximately linear on logit scale |

------------------------------------------------------------------------

## Glossary

**Accuracy:** The proportion of all predictions that were correct: (TP + TN) / Total.

**Classifier:** A model that predicts which category (class) a new observation belongs to.

**Confusion matrix:** A table comparing actual labels to predicted labels, showing true positives, true negatives, false positives, and false negatives.

**Cross-validation:** A technique for tuning model parameters by evaluating performance across multiple subsets of the training data.

**False negative (FN):** A case where the model predicted "no" but the true label was "yes" (Type II error).

**False positive (FP):** A case where the model predicted "yes" but the true label was "no" (Type I error).

**k (in k-NN):** The number of nearest neighbors used to make a classification decision. Small `k` = more flexible; large `k` = smoother.

**Log-odds (logit):** The natural log of the odds $\log(p / (1-p))$. Maps probabilities from $(0,1)$ to $(-\infty, +\infty)$.

**Null model:** A trivial baseline model that always predicts the most common class. Used to benchmark whether a learned model adds value.

**Odds:** The ratio of the probability of an event occurring to the probability of it not occurring: $p / (1-p)$.

**Odds ratio:** The ratio of two odds values, used to interpret logistic regression coefficients. An odds ratio \> 1 means the predictor increases the odds of the outcome.

**Prior probability:** The probability of a class before observing any features — $P(y)$ in Bayes' theorem.

**Test set:** Data held out from model training, used to evaluate model performance on unseen observations.

**Training set:** Data used to fit (train) a model.

**True negative (TN):** A case where the model correctly predicted "no."

**True positive (TP):** A case where the model correctly predicted "yes."
