---
title: Week 7 - Point Estimation and Confidence Intervals
author: "Valentin Milanov"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

editor_options:

## chunk_output_type: inline

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, echo = FALSE, message = FALSE}
library(ggplot2)
library(PASWR2)
library(dplyr)
library(MASS)
library(binom)
```

### Point Estimation

Review **Example 7.2** Suppose $X ~ Pois(λ)$, where $λ$ is unknown. Show
$\bar X$is an unbiased estimator of λ.

Do as an exercise! See text page 407 for details

**Solution:** Do on paper!

######################################################################################################## 

**Example 7.9** page 415 in the text.

A botanist interested in studying the effects of a new herbicide on
trifolium reopens (white clover) measures and records the stem lengths
in centimeters of ten specimens as
`5.3, 2.8, 3.4, 7.2, 8.3, 1.7, 6.2, 9.3, 3.2, and 5.9`. Compute the
mean, median, standard deviation, and MAD. Suppose the botanist makes a
field error and records an `83` instead of an `8.3`. What effect will
the recording error have on the computed quantities?

```{r ex. 7.9}

stem1 <- c(1.7, 2.8, 3.2, 3.4, 5.3, 5.9, 6.2, 7.2, 8.3, 9.3)
stem2 <- c(1.7, 2.8, 3.2, 3.4, 5.3, 5.9, 6.2, 7.2, 83, 9.3)

c(mean(stem1), sqrt(var(stem1))) # mean and std. dev. for stem 1 data
c(mean(stem2), sqrt(var(stem2)))

c(median(stem1), mad(stem1, constant = 1))  # median and MAD for stem 1 data
c(median(stem2), mad(stem2, constant = 1)) 

median(abs(stem1 - median(stem1))) # another way to compute MAD
median(abs(stem2 - median(stem2)))

```

################################################################################################# 

-   Destroyed the mean and standard deviation

-   Did not affect the median and MAD

Review **Example 8.5** page 461 in the text.

Suppose a random sample of size n from a normal distribution with
unknown mean μ and standard deviation `σ = 5` is taken. Calculate the
**minimum sample size** so that one can be `95%` confident the interval
$[\bar x-1, \bar x+1]$ contains the true value of `μ`.

REVIEW THE CODE HERE:

```{r}


nsize(b = 1, sigma = 5, conf.level = .95, type = "mu")

```

################################################################################################## 

Review **Example 8.6** page 461 in the text.

⊳Sample Size: Defective Containers⊲ In a company that produces glass
containers, the probability of producing a defective container is
$π = 0.03$, and the probability of obtaining a functional container is
$(1 − π) = 0.97$. Determine how many containers need to be manufactured
to guarantee that at least `100` containers are defective with a
probability of at least `0.95`.

**Solution:**

Let the random variable `X` again represent the number of defective
containers. The distribution of `X is`Bin(n, π = 0.03)\`. The
$P(X ≥ 100) ≥ 0.95$ is solved with the code below:

```{r}

n <- 0    # Number of containers
p <- 0    # Probability
while(p < 0.95) { # run a while cycle until p < 0.95
   n <- n + 1
   p <- 1 - pbinom(99, n, 0.03) # 1 -P(X<=99)
}

ans <- c(n, p) # answers assigned to a vector

names(ans) <- c("Containers", "Probability")
ans

```

########################################################################################################## For every 3891 containers, there will be 100 defective.

Review **Example 8.8** page 465 in the text.

Confidence Interval for `μ`: House Prices: Estimate the mean house price
for three-bedroom/two-bath houses in *Watauga County, North Carolina*. A
random sample of **14** three-bedroom/two-bath houses was taken from the
Watauga County Multiple Listing Service real estate listings (2003), and
the results are stored in the data frame `HOUSE`. Calculate a 95%
confidence interval for the average price of a three bedroom/two bath
house in this county.

**Solution:**

```{r}
data(HOUSE) # load data HOUSE
head(HOUSE) # look at the first six records

qqnorm(HOUSE$price)

xbar <- mean(HOUSE$price) # find the sample mean
CT <- qt(0.975, 13)     # critical t value
ST <- sd(HOUSE$price)   # standard deviation
xbar + c(-1, 1)*CT*ST/sqrt(14) # constructing the CI explicitly


# Second approach - quicker
t.test(HOUSE$price, conf.level = 0.95)$conf

```

**Pr. 8 / page 511 in text** The survival times in weeks for `20` male
rats that were exposed to a high level of radiation are
`152 152 115 109 137 88 94 77 160 165 125 40 128 123 136 101 62 153 83 69`.
Data are from Lawless (1982) and are stored in the data frame `RAT`.
Load the data with `data(RAT)`

(a) Construct a quantile-quantile plot of the survival times. Based on
    the quantile-quantile plot, can normality be ruled out?

(b) Construct a 92% confidence interval for the average survival time
    for male rats exposed to high levels of radiation.

**Solution:**

```{r}
data(RAT)
dim(RAT)

ggplot(data = RAT, aes(sample = survival.time)) +
  stat_qq() +
  theme_bw()

CI <- t.test(RAT$survival.time, conf.level = 0.92)$conf # Use T-distribution as we have small sample with unknown variance
CI

```

#################################################################################################### 

### 8.2.7 Confidence Interval for the Mean Difference When the Differences Have a Normal Distribution

**Example 8.17** page 481 in the text.

To compare the speed differences between two different brands of
workstations (Sun and Digital), the times each brand took to complete
complex simulations were recorded. Five complex simulations were
selected, and the five selected simulations were run on both
workstations. The resulting times in minutes for the five simulations
are given in Table 8.6 (**page text 481**) and stored in the data frame
`SUNDIG`. Construct a `95%` confidence interval for μD, the average time
difference between Sun and Digital workstations. Is one of the
workstations faster than the other?

**Solution:**

```{r}

qt(0.025, 4)
t.test(SUNDIG$sun, SUNDIG$digital, paired=TRUE)$conf

```

**Example 8.27** page 504 in the text.

**Confidence Interval and Sample Size for `π`** The Department of
Agriculture wants to estimate the proportion of rural farm owners who
are under `40` years of age. They take a random sample of 2000 farms and
find that `400` of the `2000` owners are under the age of `40`.

(a) Construct a `95%` Wald confidence interval for `π` using the
    asymptotic confidence interval for `π` based on the `MLE` of given
    in (8.43) (**text page 492**).

(b) Determine the required sample size so that the maximum margin of
    error is within `0.015` of the true value of `π` for a `95%`
    confidence level.

**Solution:**

```{r}
binom.confint(x = 400, n = 2000, conf = 0.95, methods = "asymptotic")

# if no prior information of p is available use p=1/2 this maximizes the Margin of Error 
f <- function(x){sqrt(x*(1 - x))}  # x takes place of p in f
curve(expr = f, from = 0, to = 1, n = 500) 

# use nsize() from the PASWR2 package

nsize(b = 0.015, type = "pi", conf.level = 0.95)

```

########################################################################################################### 

**Pr. 9 / page 511 in text**

A large company wants to estimate the proportion of its accounts that
are paid on time.

(a) How large a sample is needed to estimate the true proportion within
    3% with a 96% confidence level?

(b) Suppose 650 out of 800 accounts are paid on time. Construct 95%
    confidence intervals for the true proportion of accounts that are
    paid on time using an asymptotic confidence interval, an Agresti
    Coull confidence interval.

**Solution:**

```{r}
library(binom)

p <- 0.5
alpha <- 0.04
B <- 0.03

nsize(b = B, p = p, conf.level = 0.96, type = "pi") # minimum sample size

binom.confint(x = 650, n = 800, conf.level = 0.95, methods = "asymptotic")

binom.confint(x = 650, n = 800, conf.level = 0.95, methods = "ac") # Agresti Coull
```

> **Motivational Problem (Pr. 7 / page 510 in text) SOLUTION**

Surface-water salinity measurements were taken in a bottom-sampling
project in Whitewater Bay, Florida. These data are stored in the data
frame `SALINITY` in the PASWR2 package. Geographic considerations lead
geologists to believe that the salinity variation should be normally
distributed. If this is true, it means there is free mixing and
interchange between open marine water and fresh water entering the bay
(Davis, 1986).

**(a)** Construct a quantile-quantile plot of the data. Does this plot
rule out normality?

**(b)** Construct a 95% confidence interval for the mean salinity
variation.

**Solution:**

(a) The quantile-quantile plot does not rule out normality.

```{r }
ggplot(data = SALINITY, aes(sample = salinity)) +
  stat_qq() +
  theme_bw()

```

(b) 

```{r}
CI <- t.test(SALINITY$salinity, conf.level = 0.95)$conf
CI
```
